---
title: "REDCapExporter"
subtitle: "Interaction with the REDCap API"
author: "Peter DeWitt"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: true
vignette: >
 %\VignetteIndexEntry{api}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r label = "setup", include = FALSE}
# By default, chunks will not be evaluated.  This way the API token will not be
# required to build this vignette.
knitr::opts_chunk$set(collapse = TRUE, eval = FALSE)
library("REDCapExporter")
```


The purpose of this vignette is to show examples of exporting elements of a
REDCap project via the REDCap (Research Electronic Data Capture) API.  The
examples in this vignette rely on use of a API token which cannot be divulged
and thus the end users will not be able to reproduce the following examples
exactly, but hopefully will be able to use these examples as a guide for
their own use.

The raw return from the API calls has been provided and can be used as the
input for examples the end users can evaluate.

The example data provided in this package are statistics from the 2000-2001
National Hockey League Stanley Cup Champion Colorado Avalanche.  The data was
transcribed from [Hockey Reference](https://www.hockey-reference.com/teams/COL/2001.html}
into a REDCap Project hosed at the University of Colorado Denver.

This vignette is focused on the API and the general export functions.  To
learn more about automated building of a R data package from the exported
contents of a REDCap project please see


```{r eval = FALSE}
vignette(topic = "export2package", package = "REDCapExporter")
```


# REDCap API Tokens

You will need to have API export rights for the REDCap project you are
looking to export into an R data package.  Contact the project owner, system
admin, or go through your institution's REDCap webpage to acquire an API token.

Remember, your API token is the equivalent of a username/password
combination.  Thus, you must treat the token with the same, or more, level of
security you would treat any username/password combination.  **DO NOT PUT THE
TOKEN IN PLAIN TEXT!**

We suggest the reader set up a vault with their API token as a secret.
via the
`r  qwraps2::CRANpkg(secret)  `
package.  We encourage the reader to read the "secrets" vignette in R via:
`r  qwraps2::backtick(vignette(topic = "secrets", package = "secret"))  `
for details on the use of the secret package.



There are two system environment variables we recommend setting to make
the API calls simple.


```{r label = "REDCap_sysvar"}
Sys.setenv(REDCap_API_URI = "https://redcap.ucdenver.edu/api/")
Sys.setenv(REDCap_API_TOKEN = secret::get_secret("2000_2001_Avalanche"))
```


A third environmental variables is the format for data to be returned in.
Possible values for the API are 'csv', 'xml', or 'json'.   However, within
the
`r  qwraps2::CRANpkg(REDCapExporter)  `
package methods have been built to support csv or json; xml is not yet
supported.  csv is the default format.

```{r }
Sys.getenv("REDCap_API_format")
```


# Exporting a REDCap Project

First, let's load and attach the REDCapExporter namespace

```{r label = "namespace", eval = TRUE}
library(REDCapExporter)
```


Methods which will call REDCap API are
`r  qwraps2::backtick(export_content)  `
and
`r  paste0(qwraps2::backtick(export_core), ".")  `


The specific behavior and results of these functions will depended on your
institution's REDCap instance and the user access permissions associated with
the token used to access the project.

The next subsections provide details on these methods.

## Export Contents of a REDCap Project

The
`r  qwraps2::backtick(export_content)  `
method has five arguments:

```{r label = "args_of_export_content", eval = TRUE}
args(export_content)
```


* The uri, token, and format arguments are set to NULL by default.  If the
value is NULL then the system environmental variable values are used.  The
end user need only define the content argument.  Additional arguments, if
needed, are passed to RCurl::postForm via the ellipsis.

* content return specific parts of the REDCap project.

    * content = "metadata" returns the data dictionary

    * content = "record" returns the records for a project.  Note about export rights: Please be aware that Data Export user rights will be applied to this API request. For example, if you have 'No Access' data export rights in the project, then the API data export will fail and return an error. And if you have 'De-Identified' or 'Remove all tagged Identifier fields' data export rights, then some data fields *might* be removed and filtered out of the data set returned from the API. To make sure that no data is unnecessarily filtered out of your API request, you should have 'Full Data Set' export rights in the project.

    * content = "project" exports some of the basic attributes of the given REDCap project, such as the project's title, if it is longitudinal, if surveys are enabled, the time the project was created and moved to production, etc.

    * content = "user" exports the list of users for a project, including their user privileges and also email address, first name, and last name.  Note: if the user has been assigned to a user role, it will return the user with the role's defined privileges.

Check the API documentation for your host for specific additional options.
The likely uri is redcap.<institution>/api/help/.

An example: the metadata, i.e., data dictionary, for the 2000-2001 Colorado
Avalanche data set could be retrieved via

```{r }
avs_raw_metadata <- export_content(content = "metadata")
```


Since the reader does not have the API token needed to actively evaluate the
above code, the
`r  qwraps2::backtick(avs_raw_metadata)  `
object is available as a data set.


```{r label = "example_raw_metadata", eval = TRUE}
ls()
data(avs_raw_metadata)
ls()
str(avs_raw_metadata)
```


Using the as.data.frame methods will help you get the return from REDCap into
a usable form:

```{r label = "as_data_frame", eval = TRUE}
avs_metadata <- as.data.frame(avs_raw_metadata)
str(avs_metadata)
```


## Export Core

With one call to
`r  qwraps2::backtick(export_core)  `
Will call the API several times and download several elements of a REDCap
project.  The return is a list and is the expected object class to be used
as the basis for building a R data package.  An example of the return from
this method below.  It is a list of several rcer_raw_* objects.

```{r eval = TRUE}
data(avs_raw_core)
lapply(avs_raw_core, class)
```


# Session Info


```{r }
print(sessionInfo(), local = FALSE)
```

