---
title: "REDCapExporter"
subtitle: "Interaction with the REDCap API"
author: "Peter DeWitt"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: true
vignette: >
 %\VignetteIndexEntry{api}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r label = "setup", include = FALSE}
# By default, chunks will not be evaluated.  This way the API token will not be
# required to build this vignette.
knitr::opts_chunk$set(collapse = TRUE, eval = FALSE)
```



```{r label = "namespace", eval = TRUE}
library(REDCapExporter)
```


The purpose of this vignette is to show examples of exporting elements of a
REDCap project via the REDCap (Research Electronic Data Capture) API.  The
examples in this vignette rely on use of a API token which cannot be divulged
and thus the end users will not be able to reproduce the following examples
exactly, but hopefully will be able to use these examples as a guide for
their own use.

The raw return from the API calls has been provided and can be used as the
input for examples the end users can evaluate.

The example data provided in this package are statistics from the 2000-2001
National Hockey League Stanley Cup Champion Colorado Avalanche.  The data was
transcribed from [Hockey Reference](https://www.hockey-reference.com/teams/COL/2001.html}
into a REDCap Project hosed at the University of Colorado Denver.

This vignette is focused on the API and the general export functions.  To
learn more about automated building of a R data package from the exported
contents of a REDCap project please see


```{r eval = FALSE}
vignette(topic = "export", package = "REDCapExporter")
```


# System Environment Variables

`r  qwraps2::CRANpkg(REDCapExporter)  `
can use three system variables to simplify the user's interaction with the
REDCap API.  The first variable is the URI for your institution's REDCap
instance.  Set this once at the beginning of scripts, in a .Rprofile, and
this will be the default uri used in any function call with a uri argument.


```{r }
Sys.setenv(REDCap_API_URI = "https://redcap.ucdenver.edu/api/")
```


An environmental variable can be used to specify is the format for data to be
returned in from the API.
Possible values for the API are 'csv', 'xml', or 'json'.   However, within
the
`r  qwraps2::CRANpkg(REDCapExporter)  `
package methods have been built to support csv or json; xml is not yet
supported.  csv is the default format and this is set when the namespace is
loaded.

```{r }
Sys.getenv("REDCap_API_format")
```


Lastly, but the most important with respect to security, is the API token.
You will need to have API export rights for the REDCap project you are
looking to export into an R data package.  Contact the project owner, system
admin, or go through your institution's REDCap web page to acquire an API token.

Remember, your API token is the equivalent of a username/password
combination.  Thus, you must treat the token with the same, or more, level of
security you would treat any username/password combination.  **DO NOT PUT THE
TOKEN IN PLAIN TEXT!**

There are many ways for you to store, set, and use your API token, some
suggestions follow.  Consider the implications of each method carefully and
use your best judgement and care.


```{r }
Sys.setenv(REDCap_API_TOKEN = "YOURTOKENHERE")
```



# REDCap API Tokens

There are several methods to work with the token securely.  Here are some
suggestions:

## getPass

The
`r  qwraps2::CRANpkg(getPass)  `
package would let you input your token, interactively, e.g.

```{r }
Sys.setenv(REDCap_API_TOKEN = getPass::getPass())
```


The downside of this approach is that you will have to re-enter the token
every time you open the project.

## secret

The reader could set up a vault with their API token as a secret.
via the
`r  qwraps2::CRANpkg(secret)  `
package.  This package allows the user to set up a vault of secrets
encrypted via ssh keys.
We encourage the reader to read the "secrets" vignette in R via:
`r  qwraps2::backtick(vignette(topic = "secrets", package = "secret"))  `
for details on the use of the secret package.

Pros of this approach: the API key can be encrypted and stored within the
version control repository for the project you are working on.  This will
make it easy move the project form machine to machine where the same private
ssh key exists.

Downsides to this approach: if your ssh key has a passphrase you'll need to
enter it when accessing the token.  That may make non-interactive builds
difficult.  Having a specific key without a passphrase can make this approach
easier, but will require explicit setting of the ssh key (see below).  Also,
when multiple people are contributing to the project different users will
need to have specific secrets for their token(s) which will complicate the
code base.


```{r }
Sys.setenv(USER_KEY = "~/.ssh/vaults")  # ~/.ssh/id_rsa has a passphrase, ~/.ssh/vaults does not.
Sys.setenv(REDCap_API_TOKEN = secret::get_secret("2000_2001_Avalanche"))
```


Setting the environmental variable could be done via:


```{r label = "REDCap_API_sysvar"}
Sys.setenv(REDCap_API_TOKEN = secret::get_secret("2000_2001_Avalanche"))
```


## Keyring

The
`r  qwraps2::CRANpkg("keyring")  `
package can set up and access a system or custom credential store from R.

A major benefit of this approach is that one keyring could be set up for all
the API tokens you may have for dozens of individual REDCap projects.
Compared to the
`r  qwraps2::CRANpkg(secret)  `
package approach where a project vault will be needed for each project and
*might* make collaboration difficult, the keyring approach, with a little
configuration, will allow multiple users to all securely store and access
their API tokens with the same code base.

For example, the
`r  qwraps2::CRANpkg(REDCapExporter)  `
package provides a few functions to help set up a keyring, define tokens, and
access them.  In the following chunk a file based keyring called
"REDCapExporter" will be created, if needed.  The user will be prompted to
add a token for "Project1" and for "Project2."  The last line would access
the token needed for the given project.


```{r eval = FALSE}
REDCapExporter_keyring_check()
REDCapExporter_add_api_token("Project1")
REDCapExporter_add_api_token("Project2")
Sys.setenv(REDCap_API_TOKEN = REDCapExporter_get_api_token("Project1"))
```


After setting up the keyring and adding the token(s) you will only need to
use the last line setting the system environmental variable in the specific
scripts.  If the above chunk is evaluated by all collaborators on the project
with REDCap API tokens then no other modifications of the code base is needed
for the collaborators to work together.


# Exporting a REDCap Project

Methods which will call REDCap API are
`r  qwraps2::backtick(export_content)  `
and
`r  paste0(qwraps2::backtick(export_core), ".")  `

The specific behavior and results of these functions will depended on your
institution's REDCap instance and the user access permissions associated with
the token used to access the project.

The next subsections provide details on these methods.

## Export Core

With one call to
`r  qwraps2::backtick(export_core)  `
Will call the API several times and download several elements of a REDCap
project.  The return is a list and is the expected object class to be used
as the basis for building a R data package.  An example of the return from
this method below.  It is a list of several rcer_raw_* objects.

```{r eval = TRUE}
data(avs_raw_core)
lapply(avs_raw_core, class)
```


## Export Specific Content of a REDCap Project

The
`r  qwraps2::backtick(export_content)  `
method has five arguments:

```{r label = "args_of_export_content", eval = TRUE}
args(export_content)
```


* The uri, token, and format arguments are set to NULL by default.  If the
value is NULL then the system environmental variable values are used.  The
end user need only define the content argument.  Additional arguments, if
needed, are passed to RCurl::postForm via the ellipsis.

* content return specific parts of the REDCap project.

    * content = "metadata" returns the data dictionary

    * content = "record" returns the records for a project.  Note about export rights: Please be aware that Data Export user rights will be applied to this API request. For example, if you have 'No Access' data export rights in the project, then the API data export will fail and return an error. And if you have 'De-Identified' or 'Remove all tagged Identifier fields' data export rights, then some data fields *might* be removed and filtered out of the data set returned from the API. To make sure that no data is unnecessarily filtered out of your API request, you should have 'Full Data Set' export rights in the project.

    * content = "project" exports some of the basic attributes of the given REDCap project, such as the project's title, if it is longitudinal, if surveys are enabled, the time the project was created and moved to production, etc.

    * content = "user" exports the list of users for a project, including their user privileges and also email address, first name, and last name.  Note: if the user has been assigned to a user role, it will return the user with the role's defined privileges.

Check the API documentation for your host for specific additional options.
The likely uri is redcap.<institution>/api/help/.

An example: the metadata, i.e., data dictionary, for the 2000-2001 Colorado
Avalanche data set could be retrieved via

```{r }
avs_raw_metadata <- export_content(content = "metadata")
```


Since the reader does not have the API token needed to actively evaluate the
above code, the
`r  qwraps2::backtick(avs_raw_metadata)  `
object is available as a data set.


```{r label = "example_raw_metadata", eval = TRUE}
ls()
data(avs_raw_metadata)
ls()
str(avs_raw_metadata)
```


Using the as.data.frame methods will help you get the return from REDCap into
a usable form:

```{r label = "as_data_frame", eval = TRUE}
avs_metadata <- as.data.frame(avs_raw_metadata)
str(avs_metadata)
```


# Session Info


```{r }
print(sessionInfo(), local = FALSE)
```

