---
title: "REDCapExporter"
subtitle: "Building R Data Packages from REDCap Projects"
author: "Peter DeWitt"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: true
vignette: >
 %\VignetteIndexEntry{export2package}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r label = "setup", include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(REDCapExporter)
```


The purpose of this vignette is to show how to export a REDCap project into a
R data package.   Please read

```{r eval = FALSE}
vignette(topic = "api", package = "REDCapExporter")
```

for details on using the REDCap API to export the contents of a REDCap
project.  This vignette will assume you are able to call
`r  qwraps2::backtick(export_core)  `
successfully.  For the purpose of this vignette, the data set

```{r }
data(avs_raw_core, package = "REDCapExporter")
str(avs_raw_core)
```


is the result of calling
`r  qwraps2::backtick(export_core)  `
to export elements of a REDCap project containing data on the 2000-2001
Stanley Cup Champion Colorado Avalanche.

# Exporting a REDCap Project to a R Data Package

Exporting a REDCap project to a R data package is done with a call to
`r  paste0(qwraps2::backtick(build_r_data_package), ".")  `
If the reader passes the uri for the API and an API token a call to
`r  qwraps2::backtick(export_core)  `
will be made.  If you prefer, or as in this case, need to have access to the
core contents of the REDCap project available, then you may pass the
`r  qwraps2::backtick(rcer_rccore)  `
object to
`r  qwraps2::backtick(build_r_data_package)  `
to build the skeleton of a R data package.

To build the skeleton of a R data package you will need to pass in the core
export from the REDCap project, a path for were the source code for the data
package will be written, and some some information about
the users. In this context, users are the persons who have, or had, access to the
REDCap project and are listed under the UserRights section of the REDCap
project.  The user data from REDCap is used to construct the Author section
of the DESCRIPTION file for the R data package to be constructed.  By default,
all users are listed as 'contributors'.  Modification of the roles can be
provide by a named list object.  In the example below, the user dewittp is
going to assigned the creator and author role.  To be a valid R package, at
least one user will need to have the creator role assigned.


```{r }
temppath <- tempdir()
build_r_data_package(x = avs_raw_core,
                     path = temppath,
                     author_roles = list(dewittp = c("cre", "aut")))
```


The resulting directory is:

```{r }
fs::dir_tree(temppath)
```


## Details on exported Files

First, the package directory name.  Exported packages from
`r  qwraps2::Rpkg(REDCapExporter)  `
will have the directory name rcd<package-id>.  This name is also used in the
DESCRIPTION file.

The DESCRIPTION file is

```{r }
prj_dir <- list.dirs(temppath)
prj_dir <- prj_dir[grepl("/rcd\\d+$", prj_dir)]
t(read.dcf(paste(prj_dir, "DESCRIPTION", sep = "/")))
```


The title comes from the project info recorded in REDCap.  The version number
is set as the year.month.day.hour.minute of the export.  As noted above, the
Author field is built from the user data stored in REDCap.

The LICENSE file notes that the package is proprietary and should not be
installed or distributed to others who are not authorized to have access to
the data.

```{r }
cat(readLines(paste(prj_dir[1], "LICENSE", sep = "/")), sep = "\n")
```


The raw data exports are stored as .rds files under inst/raw-data so that
these files will be available in R sessions after installing the package.

The data directory has data.table versions of the data sets.

The R/datasets.R file provides the documentation for the data sets which can
be accessed in an interactive R session.


## Using the Exported Package

Let's install the package and explore the contents.

```{r }
tar_ball <- devtools::build(pkg = prj_dir)
tar_ball

install.packages(pkgs = tar_ball, lib = temppath)

```
```{r eval = FALSE}
library(rcd14465, lib.loc = temppath)

```
```{r include = FALSE}
do.call(library, args = list(package = "rcd14465", character.only = TRUE, lib.loc = temppath))
```


The available data sets:

```{r }
data(package = "rcd14465")$results
```


A simple data analysis question: how many goals were scored by position?

```{r }
library(data.table)
as.data.table(record)[, sum(goals), by = position]
```



# Session Info


```{r }
print(sessionInfo(), local = FALSE)
```

