---
title: "REDCap Exporter"
author: "Peter DeWitt"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: true
vignette: >
 %\VignetteIndexEntry{export}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r label = "setup", include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```


The purpose of this vignette is to show how to export a REDCap project into a
R data package.  There are some prerequisites, described in detail in the
first section, you will need to address before running the examples.

The primary objective of the
`r  qwraps2::Rpkg(REDCapExporter)  `
is to export all the data from an REDCap project via the REDCap API and build
a R data package with useful documentation and tools.  The data package will
make it easy to archive data and distribute data, in an analysis ready
format, within a group of analysts.

Additionally, examples and use of the specific tools from the
`r  qwraps2::Rpkg("REDCapExporter")  `
package are provided.  If you are an analyst working on a live project, the
use of these tools might be helpful for building on the fly or routine data
reports.

**Disclaimer and Warning:** It is the responsibility of the end user to
protect sensitive data.  Do not use
`r  qwraps2::Rpkg(REDCapExporter)  `
to
export data onto a computer that should not have sensitive data stored on it.
Further, do not distribute the resulting R package to any other machine or
person who does not have data access rights.

For the example in this vignette we created a REDCap project storing the
roster and statistics for the 2000-2001 Stanley Cup Champion Colorado
Avalanche of the National Hockey League.  The data was acquired from the
web page
[Hockey Reference](https://www.hockey-reference.com/teams/COL/2001.html).


# Prerequisites

## REDCap API Tokens

You will need to have API Export rights for the REDCap project you are
looking to export into an R data package.  Contact the project owner, system
admin, or go through your institution's REDCap page to acquire an API token.

Remember, your API token is the equivalent of a username/password
combination.  Thus, you must treat the token with the same, or more, level of
security you would treat any username/password combination.  **DO NOT PUT THE
TOKEN IN PLAIN TEXT!**

We suggest the reader set up a vault with their API token as a secret.
via the
`r  qwraps2::CRANpkg(secret)  `
package.  We encourage the reader to read the "secrets" vignette in R via:
`r  qwraps2::backtick(vignette(topic = "secrets", package = "secret"))  `
.

For the rest of this vignette we will use
`r  qwraps2::backtick(secret::get_secret)  `
to access a REDCap API token.

## Namespaces, Options, and System Environment Variables

For this example we will load and attach the following namespaces and set
several options and system environment variables.


```{r label = "namespacesAndOptions", message = FALSE}
library(REDCapExporter)
library(data.table)
library(magrittr)
library(secret)
library(qwraps2)

options("datatable.print.topn"  = 3)
options("datatable.print.nrows" = 6)

# Set the private key needed to access the vault, if needed.
Sys.setenv(USER_KEY = "~/.ssh/vaults")

# REDCap API uri and token
Sys.setenv(REDCap_API_uri = "https://redcap.ucdenver.edu/api/")
Sys.setenv(REDCap_API_TOKEN = get_secret("2000_2001_Avalanche"))

# Set the option the format the data will be returned in.  Possible values are
# 'csv', 'xml', or 'json'.   Methods to build data sets from csv and json have
# been built, xml is not yet supported.  csv is the default format.
Sys.getenv("REDCap_API_format")
```




# Exporting a REDCap Project to a R Data Package

## Basic Export

Exporting a REDCap project to a R data package is done with a call to
`r  qwraps2::backtick(export_redcap_project)  `
.  The output of the call is a syntactically valid source R package directory.
The arguments for the call are

```{r }
str(export_redcap_project)
```


The uri and token can be passed explicitly here, if omitted, the values in
the system environment variables will be used.
The path argument is a directory on your machine were the R data package is
going to be built.  For this vignette we will use a temporary directory.


```{r }
temppath <- tempdir()
```


Exporting the project will require you to provide at some information about
the users. In this context, users are the persons who have, or had, access to the
REDCap project and are listed under the UserRights section of the REDCap
proejct.  The user data from REDCap is used to construct the Author section
of the DESCRIPTION file for the R data package to be constucted.  By default,
all users are listed as 'contributors'.  Modification of the roles can be
provide by a named list object.  In the example below, the user dewittp is
going to assigned the creator and author role.  To be a valid R package, at
least one user will need to have the creator role assigned.



```{r }
export_redcap_project(path = temppath,
                      author_roles = list(dewittp = c("cre", "aut")))
```


The resulting directory is:

```{r }
fs::dir_tree(temppath)
```


## Details on exported Files

First, the package directory name.  Exported packages from
`r  qwraps2::Rpkg(REDCapExporter)  `
will have the directory name rcd<package-id>.  This name is also used in the
DESCRIPTION file.

The DESCRIPTION file is

```{r }
prj_dir <- list.dirs(temppath)
prj_dir <- prj_dir[grepl("/rcd\\d+$", prj_dir)]
t(read.dcf(paste(prj_dir, "DESCRIPTION", sep = "/")))
```


The title comes from the project info recorded in REDCap.  The version number
is set as the year.month.day.hour.minute of the export.  As noted above, the
Author field is built from the user data stored in REDCap.

The LICENSE file notes that the package is proprietary and should not be
installed or distributed to others who are not authorized to have access to
the data.

```{r }
cat(readLines(paste(prj_dir[1], "LICENSE", sep = "/")), sep = "\n")
```


The raw data exports are stored as .rds files under inst/raw-data so that
these files will be available in R sesssions after installing the package.

The data directory has data.table versions of the data sets.

The R/datasets.R file provides the documentation for the data sets which can
be accessed in an interactive R session.

## Other notes

### Check the R Data Package

The generated R data package should pass the basic R CMD check.

```{r }
check <- devtools::check(pkg = prj_dir, quiet = TRUE)
check
```


### Using the Exported Package

Let's install the package and explore the contents.

```{r }
install.packages(grep("*.tar.gz", list.files(path = prj_dir, recursive = TRUE, full.names = TRUE), value = TRUE),
                 lib = temppath)

library(rcd14465, lib.loc = temppath)
```


The available data sets:

```{r }
data(package = "rcd14465")$results
```


A simple data analysis question: how many goals were scored by position?

```{r }
as.data.table(record)[, sum(goals), by = position]
```




# General Tools

The
`r  qwraps2::Rpkg(REDCapExporter)  `
package provides several functions and methods which the user may find
useful.  These functions and methods are used within the package to export
and create a R data package.  Use of the methods for specific projects, say
exporting data from a REDCap project for routine reports or interim analyses.


## Specific Export Methods

The method
`r  qwraps2::backtick("export_content")  `
can be used to 
export specific parts of a REDCap project. Three of the arguments, uri,
token, and format, if omitted, will default to the system environment
variables.  Thus, only the content argument must be defined by the user.
Additional arguments for the API are passed via the ellipses.
Check your specific REDCap API documentation.


```{r }
str(export_content)
```


For example, the project information and project metadata can be obtained
via:

```{r }
avs_raw_project_info <- export_content(content = "project")
avs_raw_metadata     <- export_content(content = "metadata")
```


The return object from 
`r  qwraps2::backtick(export_content)  `
are character strings as you would
expect from a call to 
`r  qwraps2::backtick(postForm)  `
from the
`r  qwraps2::CRANpkg(RCurl)  `
package.  A couple additional attributes have been added.  The 
`r  qwraps2::backtick(Sys.time)  `
for the call and the class 
`r  qwraps2::backtick("rcer_raw_*")  `
where * is replaced by the value of the content argument.


```{r }
str(avs_raw_project_info)
str(avs_raw_metadata)
```


We opted to have 
`r  qwraps2::backtick(export_content)  `
return a character string such that the end
user can may select the format for the return and use the tool of their
choice for creating a data.frame or other objects.  However, as you will
see in the next section there are default methods for building data frames.



## Data Import Tools

When working with REDCap data it is likely that you will have more columns
than you would want to have to code a data import method for yourself.  This
is especially true if you need to deal with columns that may look like
numeric values but are really categorical.  We provide some tools to help
with the data import process.

First, you will need to get the project metadata and record.


```{r }
avs_raw_metadata <- export_content(content = "metadata")
avs_raw_metadata

avs_raw_record   <- export_content(content = "record")
avs_raw_record
```


There are 
`r  qwraps2::backtick(as.data.frame)  `
and 
`r  qwraps2::backtick(as.data.table)  `
methods for translating the raw export to the noted class.


```{r }
avs_metadata <- as.data.table(avs_raw_metadata)
avs_record   <- as.data.table(avs_raw_record)
```


The initial mapping to a data frame treats all columns as character.


```{r }
all(sapply(avs_metadata, mode) == "character")
all(sapply(avs_record, mode) == "character")
```


To create a  data frame  with useful data storage modes in each column you
can use 
`r  qwraps2::backtick(format_record)  `
.  The easiest option is to pass the record and
metadata to
`r  qwraps2::backtick(format_record)  `
as shown below.

```{r }
temp1 <- format_record(record = avs_record, metadata = avs_raw_metadata)
str(temp1[1:10])
```


The mode for each column is defined by the function call 
`r  qwraps2::backtick(col_type)  `
. The default modes are set via the metadata:


NOTES:

* Dates can be entered into REDCap as mdy, dmy, or ymd.  However, it
appears that the record will have the date in ymd format regardless of the
input format.

* Multiple Choice fields can be set as character by passing
`r  qwraps2::backtick('factor = FALSE', dequote = TRUE)  `
to 
`r  qwraps2::backtick(col_type)  `
.

* Additional defaults are to be defined.

If the user wishes to define the storage mode for a specific subset of
columns we recommend the user call
`r  qwraps2::backtick(col_type)  `
explicitly, store the result,
modify as needed, then pass the modified object to 
`r  qwraps2::backtick(format_record)  `
. For example, we will set 
`r  qwraps2::backtick(hof)  `
which is yes/no variable to logical, and modify
the height column to be in meters instead of inches.

```{r }
ct <- col_type(avs_metadata)
ct$hof <- expression(as.logical(as.integer(hof)))
ct$height <- expression(0.0254 * as.numeric(height))

temp2 <- format_record(avs_raw_record, col_type = ct)
str(temp2[1:10])
```





# Session Info


```{r }
print(sessionInfo(), local = FALSE)
#
```

